# Making maps beautiful

## Why use a map

Maps are a great way to communicate data.

They're easily understandable, flexible, and more intuitive than a chart. There's been numerous studies showing that the average reader often struggles to interpret the units on a y-axis, let alone understand trends in scatter or line graphs.

Making maps in R takes some initial investment. However once you have some code you know and understand, spinning up new pieces of analysis can happen in minutes, rather than hours or days.

The aim of this chapter is to get you from 'I can make a map in R' to something more like 'I can conduct spatial analysis and produce a visual which is ready for publication'.

## Getting started

First up, we need to load a bunch of mapping packages. The tidyverse package is a classic for just about everything data manipulation, while, the `sf` and `ggspatial` packages are essential for making maps.

```{r,results='hide', warning=FALSE,message=FALSE}

# Load required packages
library(tidyverse)  # Includes ggplot2, dplyr, tidyr, readxl, purrr
library(ggmap)  
library(sf)  
library(ggspatial)  
library(rlang)  
library(broom)  
library(Census2016)  
library(strayr)  
library(absmapsdata)  
library(officer)
```

Will Mackey's `absmapsdata` package contains all the ABS' ASGS shapefiles. The data is now also callable through the (thoroughly updated and expanded) `strayr` package. For example:

```         
strayr::read_absmap("sa12016")
```

To get a basic demographic map up and running, we will splice together the shapefile (in this case the SA2 map of Austrlaia) and some data from the 2016 Australian Census.

Hugh Parsonage put together a fantastic packaged called `Census2016` which makes downloading this data in a clean format easy.

```{r,results='hide', warning=FALSE,message=FALSE,attr.output='style="max-height: 300px;"'}
#Get the shapefile form the absmapsdata package (predefined in the list above)

#Get the 2016 census dataset
census2016_wide <- Census2016_wide_by_SA2_year

#Select the key demographic columns from the census data (i.e. the first 8 variables)
census_short <- census2016_wide[,1:8]

#Filter for a single year
census_short_2016 <- census_short %>%
    filter(year==2016)

#Use the inner_join function to get the shapefile and census wide data into a single df for analysis / visualisation
SA2_shp_census_2016 <- inner_join(strayr::read_absmap("sa22016"),census_short_2016,
                        by = c("sa2_name_2016" = "sa2_name"))

#Plot a map that uses census data 
  map1 <- ggplot() + 
          geom_sf(data = SA2_shp_census_2016, aes(fill = median_age)) +
          ggtitle("Australian median age (SA2)") +
          xlab("Longitude") +
          ylab("Latitude") + 
          theme_bw() +
          theme(legend.position = "right")
  
  map1
```

There we go! A map. This looks 'okay'... but it can be much better.

## From okay to good

Heat maps don't really show too much interesting data on such a large scale, so let's filter down to Greater Melbourne.

Seeing we have a bunch of census data in our dataframe, we can also do some basic analysis (e.g. population density).

```{r,results='hide', warning=FALSE,message=FALSE}
#As a bit of an added extra, we can create a new population density column
SA2_shp_census_2016 <- SA2_shp_census_2016 %>%
                       mutate(pop_density=persons/areasqkm_2016)

#Filter for Greater Melbourne
MEL_SA2_shp_census_2016 <- SA2_shp_census_2016 %>%
 filter(gcc_name_2016=="Greater Melbourne")

#Plot the new map just for Greater Melbourne
map2 <- ggplot() + 
          geom_sf(data = MEL_SA2_shp_census_2016, aes(fill = median_age, border=NA)) +
          ggtitle("Median age in Melbourne (SA2)") +
          xlab("Longitude") +
          ylab("Latitude") + 
          theme_bw() +
          theme(legend.position = "right")
map2
```

Much better. We can start to see some trends in this map. It looks like younger people tend to live closer to the city center. This seems logical.

## From good to great

The map above is a good start! However, how do we turn this from something 'good', into something that is 100% ready to share?

We see our 'ink to chart ratio' (i.e. the amount of non-data stuff that is on the page) is still pretty high. Is the latitude of Melbourne useful for this analysis...? Not really. Let's get rid of it and the axis labels. A few lines of code adjusting the axis, titles, and theme of the plot will go a long way. Because my geography Professor drilled it into me, I will also add a low-key scale bar.

```{r,results='hide', warning=FALSE,message=FALSE}
map3 <- ggplot() + 
          geom_sf(data = MEL_SA2_shp_census_2016, aes(fill = median_age)) +
          labs(title="Melbourne's youth tend to live closer to the city centre",
               subtitle = "Analysis from the 2016 census",
               caption = "Data: Australian Bureau of Statistics 2016",
               x="",
               y="",
               fill="Median age") + 
          ggspatial::annotation_scale(location="br")+
          theme_minimal() +
            theme(axis.ticks.x = element_blank(),axis.text.x = element_blank())+
            theme(axis.ticks.y = element_blank(),axis.text.y = element_blank())+
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            theme(legend.position = "right")+
            theme(plot.title=element_text(face="bold",size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))

map3
```

## From great to fantastic

The above is perfectly reasonable and looks professionally designed. However, this is where we can get really special.

Let's add a custom colour scheme, drop the boundary edges for the SA2's, and add in a dot and label for Melbourne CBD.

```{r,results='hide', warning=FALSE,message=FALSE}

#Add in a point for the Melbourne CBD
MEL_location <- data.frame(town_name = c("Melbourne"),
                          x = c(144.9631),
                          y = c(-37.8136))

map4 <- ggplot() + 
          geom_sf(data = MEL_SA2_shp_census_2016, aes(fill = median_age),color=NA) +
          geom_point(data=MEL_location,aes(x=x,y=y),size=2,color="black")+
          labs(title="Melbourne's youth tend to live closer to the city centre",
               subtitle = "Analysis from the 2016 census",
               caption = "Data: Australian Bureau of Statistics 2016",
               x="",
               y="",
               fill="Median age") + 
          scale_fill_steps(low="#E2E0EB", high="#3C33FE")+
          annotate(geom='curve',
                   x=144.9631,
                   y=-37.8136,
                   xend=144.9,
                   yend=-38.05,
                   curvature=0.5,
                   arrow=arrow(length=unit(2,"mm")))+
          annotate(geom='text',x=144.76,y=-38.1,label="Melbourne CBD")+
          ggspatial::annotation_scale(location="br")+
          theme_minimal() +
            theme(axis.ticks.x = element_blank(),axis.text.x = element_blank())+
            theme(axis.ticks.y = element_blank(),axis.text.y = element_blank())+
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            theme(legend.position = "right")+
            theme(plot.title=element_text(face="bold",size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))

map4

```

Now we're talking. A 'client-ready' looking map that can be added to a report, presentation, or with a few tweaks - a digital dashboard.

Make sure to export the map as a high quality PNG using the `ggplot2:ggsave()` function.
