# Australian election data

Elections tend to create fascinating data sets. They are spatial in nature, comparable over time — and more importantly they are **consequential** for all Australians.

Australia's compulsory voting system is a remarkable feature of our Federation. Every three-ish years we all turn out at over 7,000 polling booths at our local schools, churches, and community centres to cast a ballot and pick up an obligatory election day sausage. The byproduct is a fascinating longitudinal and spatial data set.

One word of warning: I use the terms electorate, division, and seat interchangeably throughout this chapter.

## Packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(strayr)
library(readaec)
library(eechidna)
```

The Australian Electoral Commission publishes detailed results for every federal election on their tally room at [results.aec.gov.au](https://results.aec.gov.au). For elections from 2007 onwards this includes first preference votes, two-party preferred, two-candidate preferred, booth-level results, polling place coordinates, Senate counts, enrolment, and turnout — all updated live on election night.

The catch is that each election has its own URL structure built around an internal event ID, and column names shift between years without warning. The [`readaec`](https://github.com/charlescoverdale/readaec) package wraps the AEC's CSV downloads in a consistent, tidy interface. One function call returns a clean data frame. Results are cached locally so you're not hitting the AEC's servers on every call.

For census analysis, the [`eechidna`](https://github.com/jforbes14/eechidna) package takes a different approach: it bundles House of Representatives results from 2001 to 2022 as ready-to-use R data frames, with ABS demographic variables already apportioned to electoral boundaries. `readaec` and `eechidna` are complementary — `readaec` for live, current election results; `eechidna` for the census join.

The [`strayr`](https://github.com/runapp-aus/strayr) package provides electoral boundary shapefiles from the ABS.

## What's available

```{r, warning=FALSE, message=FALSE}
list_elections()
```

`readaec` covers all federal elections from 2007 to 2025. The AEC doesn't provide the structured CSV downloads that `readaec` depends on for elections before 2007.

## 2025 results

Let's start with the most recent election. Who won each seat?

```{r, warning=FALSE, message=FALSE}
members_2025 <- get_members_elected(2025)
head(members_2025, 10)
```

A quick tally by party:

```{r, warning=FALSE, message=FALSE}
members_2025 |>
  count(partyab, sort = TRUE)
```

### Margins

Which seats were the most marginal?

```{r, warning=FALSE, message=FALSE}
get_tpp(2025) |>
  mutate(margin = abs(alp_pct - 50)) |>
  arrange(margin) |>
  select(division, state, alp_pct, lnp_pct, margin) |>
  head(15)
```

And the safest seats — won by the largest margins:

```{r, warning=FALSE, message=FALSE}
get_tpp(2025) |>
  mutate(margin = abs(alp_pct - 50)) |>
  arrange(desc(margin)) |>
  select(division, state, alp_pct, lnp_pct, margin) |>
  head(10)
```

## Cross-election trends

### National swing

The `get_swing()` function computes the change in two-party preferred vote between any two elections. A positive `alp_swing` means the seat moved toward Labor.

```{r, warning=FALSE, message=FALSE}
get_swing(2022, 2025) |>
  select(division, state, alp_pct_from, alp_pct_to, alp_swing, seat_changed) |>
  arrange(desc(abs(alp_swing))) |>
  head(15)
```

```{r, warning=FALSE, message=FALSE}
# How many seats changed hands?
swing_2022_25 <- get_swing(2022, 2025)
table(swing_2022_25$seat_changed)
```

### Richmond across time

Richmond (NSW) is a useful seat to trace across elections — it has a strong Greens vote that flows heavily to Labor on preferences, making it a good illustration of how preferential voting works in practice.

```{r, warning=FALSE, message=FALSE}
years <- list_elections()$year[list_elections()$has_downloads]

tpp_richmond <- map_dfr(years, function(yr) {
  get_tpp(yr) |>
    filter(tolower(division) == "richmond")
})

ggplot(tpp_richmond, aes(x = year, y = alp_pct)) +
  geom_line(colour = "#E4281B", linewidth = 1.2) +
  geom_point(colour = "#E4281B", size = 3) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 2007.2, y = 51.5,
           label = "50% — majority", size = 3, colour = "grey50", hjust = 0) +
  scale_x_continuous(breaks = years) +
  scale_y_continuous(limits = c(30, 70), labels = function(x) paste0(x, "%")) +
  labs(
    title    = "Richmond (NSW): ALP two-party preferred vote",
    subtitle = "2007 to 2025 federal elections",
    x        = NULL, y = "ALP TPP (%)",
    caption  = "Source: AEC via readaec"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Election maps

Geography matters enormously in Australian politics. Electorates vary from Durack in Western Australia (1.63 million square kilometres) to Grayndler in inner Sydney (32 square kilometres). The AEC carves up the population by state and territory, using a formula to allocate seats, then draws boundaries to achieve roughly equal populations per electorate.

At time of writing, the seat allocation is:

| **State/Territory** | **Seats** |
|---|---|
| New South Wales | 47 |
| Victoria | 39 |
| Queensland | 30 |
| Western Australia | 15 |
| South Australia | 10 |
| Tasmania | 5 |
| Australian Capital Territory | 3 |
| Northern Territory | 2 |
| **Total** | **151** |

Let's map the 2025 TPP result across all 151 divisions.

```{r, results='hide', warning=FALSE, message=FALSE}
ced2021 <- strayr::read_absmap("ced2021")
```

```{r, fig.width=7, fig.height=5, warning=FALSE, message=FALSE}
tpp_2025 <- get_tpp(2025) |>
  mutate(division = str_to_title(division))

map_data <- ced2021 |>
  rename(division = ced_name_2021) |>
  left_join(tpp_2025, by = "division")

ggplot(map_data) +
  geom_sf(aes(fill = alp_pct), colour = "white", linewidth = 0.1) +
  scale_fill_gradient2(
    low = "#80b1d3", mid = "white", high = "#E4281B",
    midpoint = 50, name = "ALP TPP %"
  ) +
  labs(
    title   = "2025 federal election: ALP two-party preferred",
    caption = "Source: AEC via readaec; boundaries: ABS 2021"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 12))
```

## Booth-level data

The AEC publishes polling place locations and vote counts at the booth level. With over 7,000 booths nationally, this is where the spatial story gets genuinely granular — even within a single electorate, booths can lean very differently.

```{r, warning=FALSE, message=FALSE}
booths <- get_polling_places(2025)
head(booths, 10)
```

```{r, fig.width=7, fig.height=4, warning=FALSE, message=FALSE}
ggplot() +
  geom_sf(data = ced2021, fill = "grey95", colour = "grey80", linewidth = 0.1) +
  geom_point(
    data = booths,
    aes(x = longitude, y = latitude),
    colour = "purple", size = 0.8, alpha = 0.3
  ) +
  xlim(112, 157) + ylim(-44, -11) +
  labs(
    title   = "Polling booths in Australia — 2025 federal election",
    caption = "Source: AEC via readaec"
  ) +
  theme_void() +
  theme(plot.title = element_text(size = 12))
```

We can also pull TPP results at the booth level. Figuring out where votes come from *within* an electorate is fundamental to campaign strategy — even small seats have pockets of strongly opposing sentiment.

```{r, warning=FALSE, message=FALSE}
tpp_booths <- get_tpp_by_booth(2025)
head(tpp_booths)
```

## Census analysis

`eechidna` does something genuinely difficult: it apportions ABS census variables to Commonwealth Electoral Division boundaries. Census data is collected at the mesh block level, but electoral boundaries don't align with census boundaries — the `eechidna` team has done the spatial concordance work so you don't have to. The result is a data frame with ~80 demographic variables, one row per electorate, ready to join directly with election results.

```{r, results='hide', warning=FALSE, message=FALSE}
data(tpp22)
data(abs2021)

election2022 <- left_join(abs2021, tpp22, by = "DivisionNm")
```

A simple but telling question: does education level predict LNP support?

```{r, warning=FALSE, message=FALSE}
ggplot(election2022, aes(x = BachelorAbv, y = LNP_Percent)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, colour = "#80b1d3") +
  labs(
    title   = "Education vs LNP two-party preferred vote — 2022",
    x       = "% with bachelor degree or above",
    y       = "LNP TPP %",
    caption = "Source: eechidna (AEC + ABS 2021)"
  ) +
  theme_minimal()
```

The negative relationship is striking — and quite recent. A university-educated vote was historically a Liberal stronghold; the teal wave of 2022 flipped that relationship in inner-city and coastal seats.

A lean regression with a handful of theoretically interesting variables:

```{r, warning=FALSE, message=FALSE}
model <- lm(
  LNP_Percent ~ BachelorAbv + MedianPersonalIncome + MedianAge + NoReligion + Indigenous,
  data = election2022
)

summary(model)
```

```{r, warning=FALSE, message=FALSE}
hist(model$residuals, col = "bisque", freq = FALSE,
     main = "Residuals", xlab = "")
lines(density(model$residuals), col = "red")
```

One caution: this is an *ecological* regression — we're modelling electorate-level aggregates, not individual votes. Interpreting the coefficients as individual-level effects would be the ecological fallacy. It tells us about the geographic pattern of voting, not about individuals.

## Informal votes

Over 700,000 people — around 5% of all votes cast — vote informally each election. Of these, over half have 'no clear first preference', meaning their vote did not contribute to the tally of any candidate.

Informal votes are genuinely fascinating. Not only are there 8 official categories (the AEC publishes a detailed breakdown [here](https://www.aec.gov.au/about_aec/research/files/analysis-of-informal-voting-2016.pdf)), but the rate of informal voting varies tremendously by electorate.

Broadly, informal votes fall into two buckets. If we want to get particular about it:

1. **Protest votes** — a person that thinks they are voting against:
   - the democratic system,
   - their local selection of candidates on the ballot, or
   - the two most likely candidates for PM.

2. **Stuff-ups** — people who:
   - filled in the form wrong but a clear preference was still made, or
   - stuffed up the form entirely and it didn't contribute towards the tally for any candidate.

The AEC works tirelessly to reduce stuff-ups on ballot papers through clear instructions and ballot design, but there isn't much of a solution for protest votes.

The geographic distribution of informal voting is worth exploring in its own right. It is not randomly distributed — the variation tells a story about disengagement and political alienation that the formal results don't capture.
