# Australian election data

Elections tend to create fascinating data sets. They are spatial in nature, comparable over time (i.e. the number of electorates roughly stays the same) - and more importantly they are **consequential** for all Australians.

Australia's compulsory voting system is a remarkable feature of our Federation. Every three-ish years we all turn out at over 7,000 polling booths our local schools, churches, and community centres to cast a ballot and pick up an obligatory election day sausage. The byproduct is a fascinating longitudinal and spatial data set.

The following code explores different R packages, election data sets, and statistical processes aimed at exploring and modelling federal elections in Australia.

One word of warning: I use the term electorates, divisions, and seats interchangeably throughout this chapter.

Let's start by loading up some packages

```{r, warning=FALSE,message=FALSE}

library(ggparliament)  # If working with parliamentary data
library(tidyverse)      # Includes dplyr, ggplot2, tidyr, purrr, and more
library(ggthemes)
library(readxl)         # Reading Excel files
library(sf)             # Spatial data handling
library(DT)
library(eechidna)
library(absmapsdata)
```

Some phenomenal Australian economists and statisticians have put together a handy election package called [eechidna](https://github.com/jforbes14/eechidna/). It includes three main data sets for the most recent Australia federal election (2022).

-   **fp22:** first preference votes for candidates at each electorate

-   **tpp22:** two party preferred votes for candidates at each electorate

-   **tcp22:** two candidate preferred votes for candidates at each electorate

They've also gone to the trouble of aggregating some census data to the electorate level. This can be found with the **abs2022** function.

```{r, warning=FALSE,message=FALSE}

data(fp22)
data(tpp22)
data(tcp22)
data(abs2022)

# Show the first few rows
head(tpp22)
head(tcp22)
```

## Election maps

As noted in the introduction, elections are ***spatial*** in nature.

Not only does geography largely determine policy decisions, we see that many electorates vote for the same party (or even the same candidate) for decades. How electorate boundaries are drawn is a long story (see [here](https://www.aph.gov.au/About_Parliament/House_of_Representatives/Powers_practice_and_procedure/Practice7/HTML/Chapter3/Electoral_divisions), [here](https://www.aec.gov.au/Electorates/Redistributions/calculating-entitlements.htm#how), and [here](https://www.aph.gov.au/About_Parliament/House_of_Representatives/Powers_practice_and_procedure/Practice7/HTML/Chapter3/Electoral_divisions)).

The summary version is the AEC carves up the population by state and territory, uses a wacky formula to decide how many seats each state and territory should be allocated, then draws maps to try and get a *roughly* equal number of people in each electorate.

Oh... and did I mention for reasons that aren't worth explaining that Tasmania has to have at least 5 seats? Our Federation is a funny thing. Anyhow, at time of writing this is how the breakdown of seats looks.

| **State/Territory** | **Number of members of the House of Representatives (2022)** |
|----|----|
| New South Wales | 47 |
| Victoria | 39 |
| Queensland | 30 |
| Western Australia | 15 |
| South Australia | 10 |
| Tasmania | 5 |
| Australian Capital Territory | 3 |
| Northern Territory | 2 |
| **TOTAL** | 151 |

*Note: The NT doesn't have the population to justify it's second seat . The AEC scheduled to dissolve it after the 2019 election but [Parliament intervened in late 2020](https://www.aec.gov.au/media/2020/12-10.htm) and a bill was passed to make sure both seats were kept (creating 151 nationally).*

How variant are these 151 electorates in size? Massive.

Durack in Western Australia (1.63 million square kilometres) is by far the largest and the smallest is Grayndler in New South Wales (32 square kilometres).

Let's make a map to make things easier.

```{r,figures-side1, fig.show="hold", out.width="50%",results='hide', warning=FALSE,message=FALSE}

CED_map <- absmapsdata::ced2021 %>%
           ggplot()+
           geom_sf()+
           labs(title="Electoral divisions in Australia",
               subtitle = "It turns out we divide the country in very non-standard blocks",
               caption = "Data: Australian Bureau of Statistics 2016",
               x="",
               y="") + 
           theme_minimal() +
            theme(axis.ticks.x = element_blank(),axis.text.x = element_blank())+
            theme(axis.ticks.y = element_blank(),axis.text.y = element_blank())+
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            theme(legend.position = "right")+
            theme(plot.title=element_text(face="bold",size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))


CED_map_remove_6 <- ced2021 %>%
                    dplyr::filter(!ced_code_2021 %in% c(506,701,404,511,321,317)) %>%   
                    ggplot()+
           geom_sf()+
           labs(title="194 electoral divisions in Australia",
               subtitle = "Turns out removing the largest 6 electorates makes a difference",
               caption = "Data: Australian Bureau of Statistics 2016",
               x="",
               y="") + 
           theme_minimal() +
            theme(axis.ticks.x = element_blank(),axis.text.x = element_blank())+
            theme(axis.ticks.y = element_blank(),axis.text.y = element_blank())+
            theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
            theme(legend.position = "right")+
            theme(plot.title=element_text(face="bold",size=12))+
            theme(plot.subtitle=element_text(size=11))+
            theme(plot.caption=element_text(size=8))

CED_map
CED_map_remove_6
```

Next let's look at what party/candidate is currently the sitting member for each electorate. To do this on a map we're going to need to join our tcp22 data and the ced2021 spatial data.

In the first data set, the electorate column in called 'DivisionNm' and in the second 'ced_name_2021'.

We see the data in our DivisionNm variable is in UPPERCASE while our ced_name_2021 variable is in Titlecase. Let's change the first variable to Titlecase to match.

We can then then join the two dataframes using the left_join function.

```{r,results='hide', warning=FALSE,message=FALSE}

#Pull in the electorate shapefiles from the absmapsdata package
electorates <- ced2021

#Make the DivisionNm Titlecase
tcp22$DivisionNm=str_to_title(tcp22$DivisionNm)

tcp22_edit <- tcp22 %>% distinct() %>% filter(Elected == "Y")

#Make the column names the same
electorates <- dplyr::rename(electorates, DivisionNm = ced_name_2021)

ced_map_data <- left_join(tcp22_edit, electorates, by = "DivisionNm")

ced_map_data <- as.data.frame(ced_map_data)

head(ced_map_data, n=151)

```

## Analysis

Let's start by answering a simple question: who won the election? For this we'll need to use the two-candidate preferred data set (to make sure we capture all the minor parties that won seats).

Note in the table above, the PartyNm variable is a mess. Some candidates noted their party by it's abbreviation rather than full name, others put in a state specific prefix For this analysis, the PartyAb variable is cleaner to use.

```{r, warning=FALSE,message=FALSE}
who_won <- tcp22 %>% 
  filter(Elected == "Y") %>% 
  group_by(PartyAb) %>% 
  tally() %>% 
  arrange(desc(n)) 

head(who_won, n=10)
```

Next up let's see which candidates won with the smallest percentage of first preference votes. Australia's preferential voting system normally makes these numbers quite interesting.

```{r,warning=FALSE,message=FALSE}

who_won_least_votes_prop <- fp22 %>% 
  filter(Elected == "Y") %>% 
  arrange(Percent) %>% 
  mutate(candidate_full_name = paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::select(candidate_full_name, PartyNm, DivisionNm, Percent)

head(who_won_least_votes_prop,n=10)

```

This is really something.

The relationship we're seeing here seems to be these are *many* the seats that are won with barely 30% of the first preference vote.

The electorate I grew up in is listed here (Richmond) - let's look at how the votes were allocated.

```{r, warning=FALSE,message=FALSE}
Richmond_fp <- fp22 %>% 
  filter(DivisionNm == "RICHMOND") %>% 
  arrange(-Percent) %>% 
  mutate(candidate_full_name = paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::select(candidate_full_name, PartyNm, DivisionNm, Percent, OrdinaryVotes)

head(Richmond_fp,n=10)
```

Sure enough - the Greens certainly helped get the ALP across the line.

The interpretation that these seats are the 'most marginal' is incorrect under a preferential voting system (e.g. imagine if ALP win 30% and the Greens win 30% - that is a pretty safe 10% margin assuming traditional preference flows).

But - let's investigate which seats are the most marginal.

```{r,warning=FALSE,message=FALSE}

who_won_smallest_margin <- tcp22 %>% 
  filter(Elected == "Y") %>% 
  mutate(percent_margin = 2*(Percent - 50), vote_margin = round(percent_margin * OrdinaryVotes / Percent)) %>% 
  arrange(Percent) %>% 
  mutate(candidate_full_name = paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::select(candidate_full_name, PartyNm, DivisionNm, Percent, OrdinaryVotes, percent_margin, vote_margin)

head(who_won_smallest_margin, n=20)
```

Crikey. We see Fiona Phillips in Gilmore got in with a 0.17% margin (just 380 votes!)

While we're at it, we better do the opposite and see who romped it in by the largest margin.

```{r,warning=FALSE,message=FALSE}

who_won_largest_margin <- tcp22 %>% 
  filter(Elected == "Y") %>% 
  mutate(percent_margin = 2*(Percent - 50), vote_margin = round(percent_margin * OrdinaryVotes / Percent)) %>% 
  arrange(desc(Percent)) %>% 
  mutate(candidate_full_name = paste0(GivenNm, " ", Surname, " (", CandidateID, ")")) %>% 
  dplyr::select(candidate_full_name, PartyNm, DivisionNm, Percent, OrdinaryVotes, percent_margin, vote_margin)

 head(who_won_largest_margin, n=20)
```

Wowza. That's really something. Some candidates won seats with a 40-45 percent margin - scooping up 70% of the two candidate preferred vote in the process!

We can also cut the seats by state for a look at where the 'strongholds' are across the country.

```{r,results='hide', warning=FALSE,message=FALSE}
who_won <- tcp22 %>% 
  filter(Elected == "Y") %>% 
  group_by(PartyAb, StateAb) %>% 
  tally() %>% 
  arrange(desc(n)) 

who_won_by_state <- spread(who_won,StateAb, n) %>% arrange(desc(NSW))

head(who_won_by_state, n=10)
```

## Trends

Now we've figured out how to work with election data - let's link it up to some AUstralian demographic data. The eechidna package includes a cleaned set of census data from 2022 that has already been adjusted from ASGS boundaries to Commonwealth Electoral Divisions.

```{r,results='hide', warning=FALSE,message=FALSE}

# Import the census data from the eechidna package
data(eechidna::abs2021)
head(abs2021)

# Join with two-party preferred voting data
data(tpp10)
election2022 <- left_join(abs2021, tpp10, by = "DivisionNm")
```

That's what we want to see. 151 rows of data (one for each electorate) and over 80 columns of census variables.

A starting exploratory exercise is too see which of these variables are correlated with voting for one party or another. There's some old narrative around LNP voters being rich, old, white, and somehow 'upper class' compared to the population at large. Let's pick a few variables that roughly match with this criteria (Income, Age, English language speakers, and Bachelor educated) and chart it compared to LNP percentage of the vote.

```{r,figures-side2, fig.show="hold", out.width="50%",results='hide', warning=FALSE,message=FALSE}

# See relationship between personal income and Liberal/National support

ggplot(election2022, aes(x = MedianPersonalIncome, y = LNP_Percent)) + 
  geom_point() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = MedianAge, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = EnglishOnly, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = BachelorAbv, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

```

First impressions: Geez this data looks messy.

Second impression: Maybe there's a bit of a trend with age and income?

Let's build a regression model to run all the 80 odd census variables in the abs2022 data set against the LNP_percent variable.

```{r,warning=FALSE,message=FALSE}

# We can use colnames(election2022) to get a big list of all the variables available

# Now we build the model
election_model <- lm(LNP_Percent~
                    Population+
                    Area+
                    Age00_04+
                    Age05_14+
                    Age15_19+
                    Age20_24+
                    Age25_34+
                    Age35_44+
                    Age45_54+
                    Age55_64+
                    Age65_74+
                    Age75_84+
                    Age85plus+
                    Anglican+
                    AusCitizen+
                    AverageHouseholdSize+
                    BachelorAbv+Born_Asia+
                    Born_MidEast+Born_SE_Europe+
                    Born_UK+
                    BornElsewhere+
                    Buddhism+
                    Catholic+
                    Christianity+
                    Couple_NoChild_House+Couple_WChild_House+
                    CurrentlyStudying+DeFacto+
                    DiffAddress+
                    DipCert+
                    EnglishOnly+
                    FamilyRatio+
                    Finance+
                    HighSchool+
                    Indigenous+
                    InternetUse+
                    Islam+
                    Judaism+
                    Laborer+
                    LFParticipation+
                    Married+
                    MedianAge+
                    MedianFamilyIncome+
                    MedianHouseholdIncome+
                    MedianLoanPay+
                    MedianPersonalIncome+
                    MedianRent+
                    Mortgage+
                    NoReligion+
                    OneParent_House+
                    Owned+
                    Professional+
                    PublicHousing+
                    Renting+
                    SocialServ+
                    SP_House+
                    Tradesperson+
                    Unemployed+
                    Volunteer,
                    data=election2022)

summary(election_model)
```

For the people that care about statistical fit and endogenous variables, you may have concerns (and rightly so) with the above approach.

It's pretty rough. Let's run a basic check to see if the residuals are normally distributed.

```{r,results='hide', warning=FALSE,message=FALSE}

hist(election_model$residuals, col="bisque", freq=FALSE, main=NA)
lines(density(election_model$residuals), col="red")
```

Hmm... that's actually not too bad. Onwards.

We see now that only a handful of these variables in the table above are statistically significant. Running an updated (and leaner) model gives:

```{r,results='hide', warning=FALSE,message=FALSE}

election_model_lean <- lm(LNP_Percent~
                          BachelorAbv+
                          CurrentlyStudying+
                          DeFacto+
                          DiffAddress+
                          Finance+HighSchool+
                          Indigenous+
                          LFParticipation+
                          Married+
                          NoReligion,
                          data=election2022)

summary(election_model_lean)
```

```{r,figures-side3, fig.show="hold", out.width="50%",results='hide', warning=FALSE,message=FALSE}

ggplot(election2022, aes(x = BachelorAbv, y = LNP_Percent)) + 
  geom_point() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = CurrentlyStudying, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = DeFacto, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = DiffAddress, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = Finance, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = HighSchool, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = Indigenous, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = LFParticipation, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = Married, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

ggplot(election2022, aes(x = NoReligion, y = LNP_Percent)) + 
  geom_jitter() + 
  geom_smooth() + 
  theme_minimal()

```

One of the more novel issues with the above analysis is that it's done on the electorate level, however electorates vary dramatically in side (google the Modifiable Area Unit Problem if this piques your interest)

Instead of using one data point for a whole electorate (regardless of how big it is), we can fetch data from all 7,000 voting booths, the match it up with the corresponding local demographic data.

## Booth data

The AEC maintains a [handy spreadsheet](http://results.aec.gov.au/20499/Website/Downloads/GeneralPollingPlacesDownload-20499.csv) of booth locations for recent federal elections. You can search for your local booth location (probably a school, church, or community center) in the table below.

```{r message=FALSE, error = TRUE, warning=FALSE, echo=FALSE, tidy=TRUE}

# Find the list of 2016 booth locations

booths <- read_csv("http://results.aec.gov.au/20499/Website/Downloads/GeneralPollingPlacesDownload-20499.csv", skip=1)

head(booths, n=10)
```

What do these booths look like on a map? Let's reuse the CED map above and plot a point for each booth location.

```{r fig.width=7, fig.height=4, message=FALSE, error = TRUE, warning=FALSE, echo=TRUE, tidy=TRUE}

ggplot() +
  geom_sf(data = ced2021) +
  geom_point(data = booths, aes(x = Longitude, y = Latitude), 
             colour = "purple", size = 1, alpha = 0.3, inherit.aes = FALSE) +
  labs(
    title = "Polling booths in Australia",
    subtitle = " ",
    caption = "Data: Australian Electoral Commission 2016",
    x = "",
    y = ""
  ) + 
  theme_minimal() +
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.title = element_text(size = 12),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 8)
  ) +
  xlim(c(112, 157)) + 
  ylim(c(-44, -11))

```

Figuring out where a candidates votes come from *within* an electorate is fundamental to developing a campaign strategy. Even in small electorates (e.g. Wentworth), there are pockets of right leaning and left leaning districts. Once you factor in preference flows, this multi-variate calculus becomes important to winning or maintaining a seat.

In the `eechidna`package, election results are provided at the resolution of polling place. Unfortunately, these are yet to be updated for elections after 2016.

The data sets must be downloaded using the functions `firstpref_pollingbooth_download`, `twoparty_pollingbooth_download` or `twocand_pollingbooth_download` (depending on the vote type).

The two files need to be merged to be useful for analysis. Both have a unique ID for the polling place that can be used to match the records. The two party preferred vote, a measure of preference between only the Australian Labor Party (ALP) and the Liberal/National Coalition (LNP), is downloaded using `twoparty_pollingbooth_download`. The preferred party is the one with the higher percentage, and we use this to colour the points indicating polling places.

We see that within some big rural electorates (e.g. in Western NSW), there are pockets of ALP preference despite the seat going to the LNP. Note that this data set is on a tpp basis - so we can't see the booths that were won by minor parties (although it would be fascinating).We see that within some big rural electorates (e.g. in Western NSW), there are pockets of ALP preference despite the seat going to the LNP. Note that this data set is on a tpp basis - so we can't see the booths that were won by minor parties (although it would be fascinating).

```{r tpp_map, fig.width=7, fig.height=5, message=FALSE, error = TRUE, warning=FALSE, echo=FALSE, tidy=TRUE}
tpp_pp <- twoparty_pollingbooth_download()
tpp_pp16 <- tpp_pp %>% filter(year == 2016)


nat_data16 <- nat_data_download(2016)
nat_map16 <- nat_map_download(2016)

# Find winner
tpp_pp16 <- tpp_pp16 %>% 
  mutate(winner = ifelse(ALP_Percent >= 50, "ALP", "LNP"))
ggplot(data=nat_data16, aes(map_id=id)) +
  geom_map(map=nat_map16, fill="grey90", colour="white") + 
  geom_point(data=tpp_pp16, aes(x=Longitude, y=Latitude, colour=winner), size=1, alpha=0.3, inherit.aes=FALSE) +
  scale_color_manual("Party", values=c("LNP"="#80b1d3", "ALP"="#fb8072")) + 
  xlim(c(112,157)) + ylim(c(-44,-11)) +
  theme_map() + coord_equal() + theme(legend.position="bottom")
```

The two candidate preferred vote (downloaded with `twocand_pollingbooth_download`) is a measure of preference between the two candidates who received the most votes through the division of preferences, where the winner has the higher percentage.

```{r tcp_map, fig.width=7, fig.height=5, message=FALSE, error = TRUE, warning=FALSE, echo=FALSE, tidy=TRUE}

tcp_pp <- twocand_pollingbooth_download()
tcp_pp16 <- tcp_pp %>% filter(year == 2016)

nat_data16 <- nat_data_download(2016)
nat_map16 <- nat_map_download(2016)

# Find winner
tcp_pp_winners <- tcp_pp16 %>% 
                  left_join(tcp_pp16 %>% dplyr::group_by(PollingPlace, DivisionNm) %>%                         summarise(TotalVotes = sum(OrdinaryVotes)), 
                            by = c("PollingPlace", "DivisionNm")) %>% 
                  filter(OrdinaryVotes/TotalVotes > 0.5)

# Plot
ggplot(data=nat_data16, aes(map_id=id)) +
  geom_map(map=nat_map16, fill="grey90", colour="white") + 
  geom_point(data=tcp_pp_winners, aes(x=Longitude, y=Latitude, colour=PartyAb), size=1, alpha=0.6, inherit.aes=FALSE) +
  scale_color_manual("Party", values=c("LNP"="#80b1d3", "ALP"="#fb8072", "GRN" = "#33a02c", "XEN" = "#beaed4", "ON" = "#fdc086", "KAP" = "#ffff99", "IND" = "grey25")) + 
  xlim(c(112,157)) + ylim(c(-44,-11)) +
  theme_map() + coord_equal() + theme(legend.position="bottom")
```

## Informal votes

Over 700,000 people (around 5% of all votes cast) vote informally each election. Of these, over have 'no clear first preference', meaning their vote did not contribute to the campaign of any candidate.

I'll be honest, informal votes absolutely *fascinate* me. Not only are there 8 types of informal votes (you can read all about the Australian Electoral Commission's analysis [here)](https://www.aec.gov.au/about_aec/research/files/analysis-of-informal-voting-2016.pdf), but the rate of informal voting varies a tremendous amount by electorate.

Broadly, we can think of informal votes in two main buckets.

1.  Protest votes

2.  Stuff-ups

If we want to get particular about it, I like to subcategorise these buckets into:

1.  Protest votes (i.e. a person that thinks they are voting against):

    -   the democratic system,

    -   their local selection of candidates on the ballot, or

    -   the two most likely candidates for PM.

2.  Stuff ups (people who):

    -   filled in the form wrong but a clear preference was still made

    -   stuffed up the form entirely and it didn't contribute towards the tally for any candidtate

The AEC works tirelessly to reduce stuff-ups on ballot papers (clear instructions and UI etc), but there isn't much of a solution for protest votes.
