# Functions

*Note: Chapter under development*

Functions are set sof instructions that take inputs (arguments), processes them, and returns an output. Instead of writing the same code repeatedly, you can define a function once and use it whenever needed. This makes your scripts cleaner and easier to maintain.

I recommend Hadley Wickham's great [chapter on functions](https://r4ds.hadley.nz/functions.html) in his book [R for Data Science 2e](https://r4ds.hadley.nz/).

## Components

A function in R is defined using the `function` keyword. Here is a simple function that adds two numbers:

```{r,results='hide', warning=FALSE,message=FALSE}

add_numbers <- function(a, b) {
  result <- a + b
  return(result)
}

# Using the function
add_numbers(3, 5)

```

In this example, we see a function consisting of:

1.  A name (e.g., `add_numbers`)
2.  Arguments (e.g., `a`, `b`)
3.  A body that performs calculations or operations
4.  A return statement (optional, but recommended)

We can also reference functions (from other packages) for use in our own custom functions. This data cleaning functions references `dplyr`, `stringr,` and `tidyr`.

```{r,results='hide', warning=FALSE,message=FALSE}

# Practical Example: Data Cleaning Function

library(dplyr)
library(stringr)
library(tidyr)

data_cleaner <- function(df) {
  df <- df %>% 
    dplyr::mutate(across(where(is.character), 
    stringr::str_trim)) %>% # Trim whitespace
    tidyr::drop_na() # Remove missing values
  return(df)
}

# Example
df <- data.frame(name = c("Alice    ", " Bob", NA), age = c(25, 30, 22))
data_cleaner(df)

```

## Luhn example

It turns out credit card numbers aren't a random 16 digits, there's an algorithm (named after it's inventor Hans Peter Luhn) that determines the combinations possible.

It follows these steps:

1.  Starting from the right, double every second digit.
2.  If doubling a digit results in a number greater than 9, subtract 9 from it.
3.  Sum all the digits.
4.  If the total sum is a multiple of 10, the number is valid; otherwise, it is invalid.

That's a lot of info... instead let's write a function to check if a credit card number is valid:

```{r,results='hide', warning=FALSE,message=FALSE}

validate_credit_card <- function(card_number) {
  digits <- as.numeric(strsplit(as.character(card_number), "")[[1]])
  n <- length(digits)
  
  # Double every second digit from the right
  for (i in seq(n - 1, 1, by = -2)) {
    digits[i] <- digits[i] * 2
    if (digits[i] > 9) {
      digits[i] <- digits[i] - 9
    }
  }
  
  # Check if the sum is a multiple of 10
  return(sum(digits) %% 10 == 0)
}

# Example usage
validate_credit_card(4532015112830366) # Should return TRUE
validate_credit_card(1234567812345678) # Should return FALSE

```

## Loops in functions

The real power of functions comes when we introduce loops. Loops allow us to repeat a calculation for different parameter values. Factorials are a great example where a loop is needed.

A *factorial* is a mathematical operation denoted by an exclamation mark (!), used to find the product of all positive integers less than or equal to a given number.

Let's write a looping function to calculate 10! (10x9x8x7x6x5x4x3x2x1)

```{r,results='hide', warning=FALSE,message=FALSE}

## Calculating Factorial Using a Loop

factorial_loop <- function(n) {
  if (n < 0) {
    stop("Factorial is not defined for negative numbers.")
  }
  result <- 1
  for (i in 1:n) {
    result <- result * i
  }
  return(result)
}

# Example usage
factorial_loop(10) # Test 10!
```

We see factorials blow out in size *very* quickly. Try inputting 100! and see what happens.
