# UK Economic Data

*Note: Chapter under development*

## ONS

The [Office of National Statistics](https://www.ons.gov.uk/) (ONS) collates and publishes most of the UK's economic data sets.

There's two main ways to access this data. Option 1, poorly formatted [spreadsheets](https://www.ons.gov.uk/economy/grossdomesticproductgdp/timeseries/abmi/pn2). Option 2, an [API](https://developer.ons.gov.uk/) (which has been in beta for over 5 years). We'll step through both.

There's several R packages out there that either ping the ONS' API or web scrape the data. Kostas Vasilopoulos has produced one of the better ones (and it's on CRAN) called `onsr`.

```{r}

library(onsr)

#Get a list of all datasets available
ons_ids()

ons_datasets()

onsgdp <- onsr::ons_get(id = "gdp-to-four-decimal-places")

head(onsgdp,100)

```

## Nomis

Census data, as well as economic and social data sets that take into account location, are produced by the ONS' [Nomis](https://www.nomisweb.co.uk/) service. Nomis has it's [very own R package](https://github.com/ropensci/nomisr) called `nomisr`.

The last UK census was in 2021. While there is a census slated for 2031, there is a [growing consensus](https://www.bbc.co.uk/news/uk-51468919) that this may not be necessary, with new administrative data sets providing a more accurate picture of the UK's population.

```{r}

```

## Trade

In April 2025, trade is well and truly the talk of the town.

UK trade data is collated by HM Revenue & Customs (HMRC) and published on the website <https://www.uktradeinfo.com/>.

[Peter van der Meulen](https://github.com/pvdmeulen) has developed a wonderful R package [`uktrade`](https://github.com/pvdmeulen/uktrade) to query HMRC's database and stitch together tables.

Here's a worked example:

```{r}

```
