# UK Economic Data

The UK has a well-developed statistical infrastructure, and most of it is freely accessible via R. This chapter covers three complementary packages that between them cover the main sources: the Office for National Statistics, the Nomis labour market database, and the Bank of England.

## Packages

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(onsr)
library(nomisdata)
library(bbk)
```

A quick map of what each package covers:

- **`onsr`** — R client for the [ONS API](https://developer.ons.gov.uk/), giving access to national accounts, GDP, inflation, and hundreds of other official statistics
- **`nomisdata`** — Access to [Nomis](https://www.nomisweb.co.uk/), the ONS service for labour market and census data broken down by geography
- **`bbk`** — Client for the [Bank of England's statistical database](https://www.bankofengland.co.uk/boeapps/database), covering interest rates, money supply, exchange rates, and more

All three are on CRAN and require no API key for standard use.

## ONS: Regional GDP

The ONS publishes national accounts data through its API. The `ons_get()` function takes a dataset ID and returns a tidy data frame. Use `ons_ids()` to see all available datasets.

```{r, warning=FALSE, message=FALSE}
gdp <- ons_get(id = "regional-gdp-by-year")
head(gdp)
```

The dataset is multi-dimensional — broken down by region, industry, price measure, and series type. We filter to total industry output (`A--T`) and the annual index (2019 = 100), which lets us compare how different UK regions have grown over time.

```{r, warning=FALSE, message=FALSE}
gdp_regional <- gdp |>
  filter(`sic-unofficial` == "A--T",
         GrowthRate == "Annual index",
         !Geography %in% c("Extra-regio")) |>
  arrange(Geography, Time)

ggplot(gdp_regional, aes(x = Time, y = v4_1, colour = Geography)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 2012.1, y = 101.5, label = "2019 = 100",
           size = 3, colour = "grey50", hjust = 0) +
  labs(
    title   = "UK regional GDP index",
    subtitle = "Chained volume measure, 2019 = 100",
    x       = NULL,
    y       = "Index",
    colour  = NULL,
    caption = "Source: ONS via onsr"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        legend.position = "right")
```

## Nomis: Labour market

Nomis is the ONS's specialist service for labour market and census data, with results broken down to local authority and ward level. The `fetch_nomis()` function takes a dataset ID and filter parameters.

```{r, warning=FALSE, message=FALSE}
# UK unemployment by region
unemployment <- fetch_nomis(
  id       = "NM_1_1",
  time     = "latest",
  geography = "TYPE480",
  measures = 20100,
  sex      = 7
)

head(unemployment)
```

```{r, warning=FALSE, message=FALSE}
unemployment |>
  filter(!is.na(OBS_VALUE), ITEM_NAME == "Total claimants") |>
  arrange(desc(OBS_VALUE)) |>
  ggplot(aes(x = reorder(GEOGRAPHY_NAME, OBS_VALUE), y = OBS_VALUE)) +
  geom_col(fill = "#1f77b4") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title   = "UK unemployment claimants by region",
    subtitle = paste("JSA claimants —", format(Sys.Date(), "%B %Y")),
    x       = NULL,
    y       = "Claimant count",
    caption = "Source: Nomis via nomisdata"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

## Bank of England: Base rate

The `bbk` package covers several central banks — the `boe_*` functions are specific to the Bank of England. The series ID `IUMABEDR` is the official Bank Rate.

```{r, warning=FALSE, message=FALSE}
bank_rate <- boe_data(
  key        = "IUMABEDR",
  start_date = "2000-01-01"
)

head(bank_rate)
```

```{r, warning=FALSE, message=FALSE}
bank_rate |>
  filter(!is.na(value)) |>
  ggplot(aes(x = date, y = value)) +
  geom_line(colour = "#d62728", linewidth = 0.9) +
  labs(
    title   = "Bank of England official bank rate",
    subtitle = "2000 to present",
    x       = NULL,
    y       = "Rate (%)",
    caption = "Source: Bank of England via bbk"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The post-2022 rate cycle is striking in context — rates went from near-zero (where they had sat since the 2008 financial crisis) to over 5% in the space of two years, before starting to come back down.
